Assessing linkages between watershed nutrient loading and estuary water quality in Lavaca Bay, Texas

### 2023-02-10 Schramm-Manuscript-2023/Schramm-Manuscript-2023.tex
Intro kinda done

% Simple summary
\simplesumm{A Simple summary goes here.}

@@

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
keywords specific to the article, yet reasonably common within the
subject discipline.).}

@@

Like many estuaries globally, estuaries along the Texas Gulf coast are
facing pressures from increasing population, increases in point source
and non-point source pollution and alterations to freshwater inflows
leading to increases in the occurrences and risks of algal blooms and
eutrophication
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Recent studies indicate that estuary water quality dynamics in both
agriculturally dominated and urban watersheds within Texas are
displaying signals of conditions increasingly conducive to
eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.

@@

\begin{figure}[H]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{Schramm-Manuscript-2023_files/figure-latex/fig1-1.png}
\end{adjustwidth}
\caption{Map of the Lavaca Bay watershed, location of USGS gages where nutrient loads were calculated, and location of estuary water quality sampling sites.\label{fig1}}
\end{figure}

Daily discharges for the Lavaca River (USGS-08164000) were obtained from
the United States Geologic Survey (USGS) National Water Information
System using the \emph{dataRetrieval} R package
\citep{deciccoDataRetrievalPackagesDiscovering2022}. Gaged dailly
discharges from Lake Texana (USGS-0816425) were provided by the Texas
Water Development Board (TWDB) (April 21, 2022 email from R. Neupane,
TWDB).

Water quality sample data for both freshwater and estuary locations were
o btained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. The QAPP and procedures manuals ensure the consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. For freshwater
locations, total phosphorus (TP) and nitrate-nitrogen
(NO\textsubscript{3}) data were downloaded (Table \ref{tab:fwsummary}).
Unfortunately, insufficient data was available for assessment of total
nitrogen (TN) and total Kjeldahl nitrogen (TKN) loadings, so analysis
was restricted to TP and NO\textsubscript{3} loads. For estuary
locations, we obtained data for TP, Nitrite+Nitrate
(NO\emph{\textsubscript{x}}), TKN, chlorophyll-\emph{a}, and dissolved
oxygen (Table \ref{tab:estuarysummary}).

@@

Estimates of nutrient loads were developed using Generalized Additive
Models (GAMs) relating nutrient concentration to river discharge,
season, and time. Separate models were fit at each station for each
parameter and used to predict nutrient concentrations for each day in
the study period. GAMs can be specified in a functionally similar manner
to the commonly used LOADEST \citep{cohn_validity_1992} or WRTDS
\citep{hirsch_weighted_2010} regression models and have been shown to
produce reliable estimates of nutrient and sediment loadings
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnert_quantifying_2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022}.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{wood_fast_2011}. Although the underlying parameter estimation
procedure of GAMs is substantially different than WRTDS, both the
functional form and results are demonstrated to be similar
\citep{beckNumericalQualitativeContrasts2017}. The use of GAMs over
other regression-based approaches was (1) the ability to easily explore
and incorporate different model terms, (2) the ability to incorporate
non-linear smooth function without explicit apriori knowledge of the
expect shape, and (3) the ability to specify a link function that
relates the expected value of the response to the linear predictors and
allows use to avoid data transformations as much as possible.

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period.
\citet{wangLoadEstimationUncertainties2011},
\citet{kuhnert_quantifying_2012} and \citet{zhang_improving_2017} refer
to this as averaged or smoothed discounted flow and demonstrated
improvements in nutrient loading models by including the term.
\citet{kuhnert_quantifying_2012} expresses MA as:

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' based on median NSE,
r\textsuperscript{2} and PBIAS calculated from repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}).In
addition to higher average NSE and r\textsuperscript{2} values, GAMs had
smaller variance in metric values for TP compared to
NO\textsubscript{3}.

\begin{figure}[H]\begin{adjustwidth}{-\extralength}{0cm}

@@

\begin{figure}[H]

@@

\caption[Aggregated estimated annual and flow-normalized annual NO  extsubscript{3} and TP loads for USGS-08164000 and USGS-08164525]{Aggregated estimated annual and flow-normalized annual NO extsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.}\label{fig:fig3}
\end{figure}

\begin{figure}[H]

@@

\begin{figure}[H]\begin{adjustwidth}{-\extralength}{0cm}

@@

@@

Authors should discuss the results and how they can be interpreted in
perspective of previous studies and of the working hypotheses. The
findings and their implications should be discussed in the broadest
context possible. Future research directions may also be highlighted.

@@

\acknowledgments{All sources of funding of the study should be
disclosed. Please clearly indicate grants that you have received in
support of your research work. Clearly state if you received funds for
covering the costs to publish in open access.}

\conflictsofinterest{The author declares no conflict of interest. The
founding sponsors had no role in the design of the study; in the
collection, analyses, or interpretation of data; in the writing of the
manuscript, an in the decision to publish the results.}

>>>>

% Simple summary

@@

% Keywords
\keyword{estuary; nutrient loading; water quality; Texas}

@@

Like many coastal areas globally, coastal watersheds along the Texas
Gulf coast are facing pressures from increasing population, increases in
point source and non-point source pollution and alterations to
freshwater flows that alter water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these increasing pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or lower for
exhibiting eutrophic conditions \citep{bricker_effects_2008}. However, a
suite of recent studies indicate that estuary water quality dynamics in
both agriculturally dominated and urban watersheds within Texas are in
fact expressing conditions that are increasingly conducive to algal
blooms and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of localized areas of estuary water quality concern
along the Texas coast \citep{bugica_water_2020}, localized studies are
being prioritized to better inform management actions.

This project provides an assessment of nutrient loading and water
quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in
the larger Matagorda Bay system located roughly halfway between Houston,
Texas and Corpus Christi, Texas. Lavaca Bay has faced substantial
challenges associated with legacy contamination but general water
quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. However, long-term declines of benthic fauna abundance,
biomass, and diversity in Lavaca Bay that are linked to reductions in
freshwater inflows and changes in estuary salinity
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}
and are a concern to local stakeholders. \citet{bugica_water_2020}
identified monotonic increases in total phosphorus (TP), orthophosphate,
total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a} at sites within
Lavaca Bay. Although no long-term changes in DO concentrations were
identified, the trends in nutrient concentrations are concerning due to
the role of nitrogen as a limiting factor for primary production in many
Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that major changes in nitrogen loadings could have
for productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schramm_lavaca_2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to better
identify and replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
To provide actionable information for resource managers, water quality
conditions must be evaluated relative to changes in natural
environmental drivers to better understand and manage potential
anthropogenic effects.

Here something some we present methodology and results for estimating
nutrient loading- and assessing linkages in estuary water quality
parameters with watershed derived nutrient loads\ldots.

@@

\textbf{Need to add a short description of land use}

Daily discharges for the Lavaca River (USGS-08164000, Figure
\ref{fig:fig1}) were obtained from the United States Geologic Survey
(USGS) National Water Information System using the \emph{dataRetrieval}
R package \citep{deciccoDataRetrievalPackagesDiscovering2022}. Gaged
daily discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

Water quality sample data for the two freshwater and three estuary
locations were obtained from the Texas Commission on Environmental
Quality (TCEQ) Surface Water Quality Monitoring Information System. Data
submitted through the system are required to be collected under Quality
Assurance Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. The QAPP and procedures manuals ensure the consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying lengths of and availability of data. For freshwater locations,
TP from January 2000 through December 2020 and nitrate-nitrogen
(NO\textsubscript{3}) data from January 2005 through December 2020 were
downloaded (Table \ref{tab:fwsummary}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater
sites and deemed insufficient to develop load estimation models
(\textbf{CITE}). The three estuary sites included an upper Lavaca Bay
site near the outlet of the Lavaca River system (TCEQ-13563), a
mid-Lavaca Bay site (TCEQ-13383), and the lower Lavaca Bay site near the
mouth of the Bay (TCEQ-13384). For estuary locations, we obtained data
for TP, Nitrite+Nitrate (NO\emph{\textsubscript{x}}), TKN,
chlorophyll-\emph{a}, and DO concentrations from January 2005 through
December 2020 (Table \ref{tab:estuarysummary}).

@@

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164250) were developed using Generalized Additive Models (GAMs)
relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to
predict nutrient concentrations for each day in the study period. GAMs
can be specified in a functionally similar manner to the commonly used
LOADEST \citep{cohn_validity_1992} or WRTDS \citep{hirsch_weighted_2010}
regression models and have been shown to produce reliable estimates of
nutrient and sediment loadings
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnert_quantifying_2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022}.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{wood_fast_2011}. Although the underlying parameter estimation
procedure of GAMs is substantially different than WRTDS, both the
functional form and results are demonstrated to be similar
\citep{beckNumericalQualitativeContrasts2017}. The use of GAMs over
other regression-based approaches was (1) the ability to easily explore
and incorporate different model terms, (2) the ability to incorporate
non-linear smooth function without explicit apriori knowledge of the
expect shape, and (3) the ability to specify a link function that
relates the expected value of the response to the linear predictors and
allows use to avoid data transformations as much as possible.

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period.
\citet{wangLoadEstimationUncertainties2011},
\citet{kuhnert_quantifying_2012} and \citet{zhang_improving_2017} refer
to this as averaged or smoothed discounted flow and demonstrated
improvements in nutrient loading models by including the term.
\citet{kuhnert_quantifying_2012} expresses \emph{ma} as:

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' using median NSE,
r\textsuperscript{2} and PBIAS calculated using repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}).In
addition to higher average NSE and r\textsuperscript{2} values, GAMs had
smaller variance in metric values for TP compared to
NO\textsubscript{3}.

\begin{figure}\begin{adjustwidth}{-\extralength}{0cm}

@@

\begin{figure}

@@

\caption[Aggregated estimated annual and flow-normalized annual NO\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525]{Aggregated estimated annual and flow-normalized annual NO\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.}\label{fig:fig3}
\end{figure}

\begin{figure}

@@

\begin{figure}\begin{adjustwidth}{-\extralength}{0cm}

@@

TP and NO\textsubscript{3} loadings from the Lavaca Bay watershed showed
high inter-annual variability tied with changes in discharge. There is
little evidence for changes in flow-normalized TP loads in either
rivers. There is some evidence of recent decreases in flow-normalized
NO\textsubscript{3} loads in the Lavaca River. Although there is no work
directly correlating water quality planning and implementation efforts
in the watershed to water quality outcomes, efforts to increase
agricultural producer participation in the watershed have been ongoing
since 2016
\citep{schramm_lavaca_2018, bertholdDirectMailingEducation2021}. The
decrease in flow-normalized NO\textsubscript{3} loads could be a
reflection of those collective efforts but further data collection and
research is required to support that statement.

Other studies have published estimates of TP loading estimates for the
Lavaca River (USGS-08164000) over various time periods (Table
\ref{tab:tpcomp}). Converted to average annual yield, the estimates of
TP loading in this study are within the ranges of previous studies
\citep{dunnTrendsNutrientInflows1996, rebich_sources_2011, omaniEstimationSedimentNutrient2014, wise_spatially_2019},
with the TP estimates in \citet{dunnTrendsNutrientInflows1996} being
notably lower. Given that none of the studies identify substantially
sized trends in TP, it is is possible that the period used in
\citet{dunnTrendsNutrientInflows1996} was drier on average than the
other studies. The SPARROW models used in \citet{rebich_sources_2011}
and \citet{wise_spatially_2019} utilize a version of LOADEST in the
underlying load estimation procedure, so a difference due to methodology
alone is unlikely.

@@

Cross-validation of the GAM loading models indicated that GAMs performed
well on average at predicting daily nutrient loading values. The
variance in scores was very high indicating subsets of values were
problematic at characterizing functional relationships between nutrients
and predictors. All of the water quality data for the two river sites in
the TCEQ database were ambient water quality data, collected to be
representative of typical conditions. This results in few data collected
at the highest portions of the flow duration curve under which the
majority of loadings take place. Supplementary flow-biased monitoring
targeting storm- or high-flow conditions is recommended here to improve
the precision of GAM predictions
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.

The non-linear temporal water quality trends identified using GAMs
differed slightly from the trends identified by
\citet{bugicaWaterQualityTrends2020}. This is not unexpected due to the
different time periods, different methodology, and generally small
slopes identified for most of the significant water quality parameters
in \citet{bugica_water_2020}. The trend in DO and cholorophyll-\emph{a}
concentrations are stable in comparison to other Texas estuaries that
might be facing larger demands for freshwater diversions, higher
population growth, and more intense agricultural production
\citep{wetzWaterQualityDynamics2016, bugica_water_2020}. The trend of
increasing NO\textsubscript{\emph{x}} concentration at the upper-bay
TCEQ-13563 site and recent increases in TKN concentration at the mid-bay
TCEQ-13383 site are concerning due to the nitrogen limitation identified
in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{x}}, TKN, and TP are suggestive of nonpoint
watershed sources, consistent with watershed uses, and consistent with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\citep{russell_effect_2006, caffreyHighNutrientPulses2007, peierlsNonmonotonicResponsesPhytoplankton2012, palmerImpactsDroughtsLow2015, ciraPhytoplanktonDynamicsLowInflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. Tidal flushing
from Matagorda Bay might be responsible for decreasing TKN
concentrations and limiting the effects of freshwater inflow in lower
reaches of Lavaca Bay. \citet{russell_effect_2006} suggested the
processing of organic loads in the upper portions of Lavaca Bay reduces
the transport of nutrients into the lower reaches of the Bay.

Freshwater inflow also had a strong positive effect on
chlorophyll-\emph{a} at the upper- and mid-bay sites. The upper-bay
site, TCEQ-13563, showed decreases in chlorophyll-\emph{a} at the
highest freshwater inflow volumes. Freshwater flushing or increases in
turbidity are associated with decreases in chlorophyll-\emph{a} in other
estuaries
\citep{peierlsNonmonotonicResponsesPhytoplankton2012, cloernPhytoplanktonPrimaryProduction2014}.
No relationships between inorganic nitrogen or TP loadings with
chlorophyll-\emph{a} were observed. Due to the lack of TKN loading
information, no assessment between organic nitrogen loads and
chlorophyll-\emph{a} were possible.

Although other studies have identified complex relationships between
estuary nutrient concentrations, nutrient loading and
chlorophyll-\emph{a} concentrations in Texas estuaries
\citep{ornolfsdottirNutrientPulsingRegulator2004, doradoUnderstandingInteractionsFreshwater2015, ciraPhytoplanktonDynamicsLowInflow2021, tominackVariabilityPhytoplanktonBiomass2022},
this study specifically used flow-adjusted freshwater derived nutrient
loads to parse out contributions from changes in nutrient loadings while
accounting for variations in load due to flow. Loading GAMs indicated no
evidence of changes in flow-normalized TP loads in either river, and no
changes in flow-normalized NO\textsubscript{3} loads in the Navidad
River. The small changes in flow-normalized NO\textsubscript{3} loads in
the Lavaca River are probably masked under most conditions by discharge
from the Navidad River. Given the relatively little variation in
flow-normalized loads, it can be expected that they would contribute
little to the variance in downstream water quality.

GAMs did not identify responses in DO concentration to inflows or
nutrient loads. The seasonality term in the temporal GAM models
explained a substantial amount of DO variation at all of the sites.
Responses of estuary metabolic processes and resulting DO concentrations
can be quite complicated and often locally specific
\citep{caffreyFactorsControllingNet2004}. While the lack of total
nitrogen or TKN loading data hinders interpretation, the large seasonal
effect on DO suggests physical factors play an important role and should
be included in future models. Prior work suggests that there is some
evidence the Lavaca Bay may not be limited by nutrients alone, with high
turbidity or nutrient processing in upper portions of the Bay or
intertidal river limiting production \citep{russell_effect_2006}.

@@

\acknowledgments{Thank you to Mike Wetz (Harte Institute, Texas A\&M
Corpus Christi), Chad Kinsfather and Partick Brzozowski (Lavaca-Navidad
River Authority), Brian Koch (Texas State Soil and Water Conservation
Board), Bill Balboa (Matagorda Bay Foundation) and the Lavaca Bay
Foundation for supporting development of this project and providing
valuable feedback. Thank you to Ram Neupane (Texas Water Development
Board) for providing data.}

\conflictsofinterest{The author declares no conflict of interest. The
project sponsors had no role in the design of the study; in the
collection, analyses, or interpretation of data; in the writing of the
manuscript, or in the decision to publish the results.}
